Relationships in the extremes and their influence on competition and coexistence

### 2022-04-21 paper/Paper.Rnw
finished editing the Discussion

\author[a]{Pimsupa Jasmin Albert}
\author[a,b,*]{Daniel C. Reuman}

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas}
\affil[b]{Laboratory of Populations, Rockefeller University}
\affil[*]{Corresponding author, reuman@ku.edu}

@@

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using the lottery model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
%***DAN: Make sure we end up having such material in the Discussion
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical and empirical importance.

@@

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a strong competitor is present across a
network of habitat patches and the GWR of a weaker competitor is only positive 
in a patch in the presence of 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present,
and can therefore be regarded as an ``ATA specialist'' in the landscape. EC-associations
can often boil down to associations between two environmental variables, as was the case
for the lottery model we considered. Mechanisms of coexitence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs are absent, and is hence a specialist on scenarios with symmetric
tail associations of environmental variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
In the Introduction we explained a plausible means by which climate change may
alter the tail association properties of phenologically separated environmental variables
such as early- and late-spring rainfall if, as seems likely, meteological events which
are more extreme also tend to be longer lasting. The possibility 
that changes in the nature of environmental ATAs 
may result in the extirpation of ATA and STA specialists indicates that it may
be important to engage in further study of ATAs between environmental variables 
and how they differ across space and may be changing through time due to human
impacts on the climate. Extreme environmental events have been much studied, including
how the nature of such events may be changing with climate change, and how such changes 
may influence species coexistence (see citations in the Introduction). But
studies of changes in environmental ATAs would differ from such prior work because
ATAs pertain to relationships between variables in the extremes, a potential changes in those
relationships, to the best of our knowledge, have not yet been examined.

%E and vital rates
Whereas $E$ of the $EC$ associatons which produce storage effects 
is often conceptualized as ``environment'', in fact it is a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is actually better conceptualized as a species response
than as the environment, directly.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, clearly a response
to the environment rather than a direct measurement of the environment. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
Thus the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population quantities which actually matter for species coexistence.
Vital rates are often measured directly in demongraphic studies. To our knowledge, the 
nature of tail associations between vital rates has been studied seldom or never.

@@

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms result from rendering one or more of $X$, $Y$, $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present some difficulties.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.
It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}.

%***DAN: stopped here 2022 04 21

%\underline{caveats and future directions}
%Realism - we usede simple/extreme ATA, simple model and simple lab system, futurw work needed to assess 
%how mcuh ATA are important in complex systems, field systems
%what systems? what env vars?
%Ellner et al (2016) points out that coexistence is also mediated by other factors that competition: shared enemies, 
%mutualist, facilitation etc. (these are the exact factors Ellner lists), so future work could explore 
%the importance of ATA for coexistence mediated by these factors.
%Something about spatial mechanisms of coexistence? We considered only mechanisms that applied in a model with
%solely temporal variation, but there are also spatial mechanisms. Need to learn about those and think about
%how to say something about how future work may be able to explore the importance of ATA in that context?

%Also need to talk about how this stuff will apply to data.
As an introductory investigation of ATAs within the framework of MCT, 
we used simple and extreme examples, and future work should assess more complex relationships, systems, and models. 
In our theoretical exploration, we implemented a simple lottery model and assessed 
the influence of extremely asymmetric associated noise, i.e., the noise was perfectly correlated
in exactly one half of the distribution. However, more complex models (such as structured demographic models
or integral projection models should be explored )and varying degrees and qualities of ATA, (copula parameters, 
difference between tails, rate of relatedness(slope of expansion??) should be implemented and explored.
Moreover, because our empirical application makes use of a simple and controlled lab system, future studies
should be done to assess the extent of how much ATAs are important for complex systems and field systems which are 
often subject to factors beyond our control.
Because modeling the dynamics of long-lived populations can be challenging, a good field system to start
exploring the ideas presented here is a system of annual plants. 
For example, the ATAs effects of rainfall timing on the coexistence of grasses and forbs 
in California grasslands \citep{Hallett_2019} could be a fruitful 
further investigation of biodiversity maintenance in California grasslands and the effects of extreme 
rainfall fluctuation given the highly variable climate of the area.
Moreover, our approach may provide more insight when expanded to consider fluctuations across space,
in regards to questions about population synchrony, Moran effects, and the suggestion of possible ATA specialists.

%Delta=1 means no overlapping generations -  breaking the storage effect requirement of buffered population growth.

%\underline{conclude **NOT NEEDED**: how ur work answers/informs the broad question from the intro and lay out 2-3 key priorities for next research on this question}
%We provided another showcase of the importance of ATAs in ecology, 
%building on previous studies.
%ATAs clearly have implications for coexistence and competition;
%though more research is needed in exploring more complex models 
%including spatiotemporal and multispecies models.
%Hope to see this tertiary decomposition of MCT be applied to empirical research.
%More investigations should be done to explore implications of ATAs 
%to other ecological theories.

%Search for comments in the rest of the text indicating things that need to be done in the Discussion

%Knowledge of stochastic structure (Shoemaker et al 2020)

@@

The authors thank Shyamolina Ghosh for suggesting the research problem, and Lawrence Sheppard, Max Castorani,
Jon Walter, Jude Kastens, Vadim Karatayev, Emily Arsenault, Neil Thimmesch, Steve Ellner, and Robin Snyder for helpful discussions. 
The authors were partly supported by US National
Science Foundation grants 1714195 and 2023474, the James S McDonnell Foundation, and the California Department of Fish and
Wildlife Delta Science Program. 
%***DAN: Jasmim, please add any other acknolwedgments here - anyone else you have discussed this with or anything like that.
%Be generous here, add people even if they only helped a tiny bit. If Steve consents to be coauthor, we take him
%off this list and add him to the author list.
%***DAN: DAN: if we add Ellner to the author list take him out here, and also take him out of the acks in the README.

@@

%***DAN: Jasmin, some changes to this figure are needed. Please make axis labels and numbers bigger. Please put 
%axis labels outside the axes instead of inside. Please make panel labels bigger. Please make coutour labels
%slightly bigger.
%***DAN: Do we want to do anything with standard error for this fig?
\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
the GWR itself. The ratio $\Delta_i^{[EC]}/\text{GWR}$ is shown, larger magnitudes
indicating greater importance of ATAs and values greater than $1$ corresponding to either
ATA rescue or ATA exclusion (Methods). For this system and the plotted parameters, 
$\Delta_i^{[EC]}$ was always positive, so cyan indicates exclusion of \emph{F. crotonensis} and
magenta indicates coexistence of the two diatom species. Values greater than 1 or less than -1
were remapped to 1 and -1, respectively, for readability.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, was always positive both including and
excluding ATAs, so GWR for \emph{F. crotonensis} is a sufficient coexistence metric.
Standard errors of plotted quantities were always less then \Sexpr{round(se6,5)}, so are not shown.} \label{fig:fig6}
\end{figure}

>>>>

\author[a]{Pimsupa Jasmin Albert}
\author[b]{Stephen P. Ellner}
\author[a,c,*]{Daniel C. Reuman}

\affil[a]{Department of Ecology and Evolutionary Biology and Center for Ecological Research, University of Kansas}
\affil[b]{Department of Ecology and Evolutionary Biology, Cornell University}
\affil[c]{Laboratory of Populations, Rockefeller University}
\affil[*]{Corresponding author, reuman@ku.edu}

@@

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using the lottery model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical and empirical importance.

@@

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a strong competitor is present across a
collection of habitat patches and the GWR of a weaker competitor is only positive 
in a patch in the presence of 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present.
The weaker competitor can therefore perhaps be regarded as an ``ATA specialist'' in the landscape. EC-associations
can often boil down to associations between two environmental variables, as was assumed to be the case
for the lottery model. Mechanisms of coexitence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs are absent, and is hence a specialist on scenarios with symmetric
tail associations of environmental variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
In the Introduction we explained a plausible means by which climate change may
alter the tail association properties of phenologically separated environmental variables
such as early- and late-spring rainfall. The possibility 
that changes in the nature of environmental ATAs 
may result in the extirpation of ATA and STA specialists indicates that it may
be important to engage in further study of ATAs between environmental variables 
and how they differ across space and may be changing through time. 
Extreme environmental events have been much studied, including
how the strength and commonness of such events may be changing with climate change, and how such changes 
may influence coexistence (see citations in the Introduction). But
studies of changes in environmental ATAs would differ from such prior work because
ATAs pertain to \emph{relationships} between variables in the extremes.

%E and vital rates
Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', in fact it is a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is actually better conceptualized as a species response
than as the environment, directly.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, clearly a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
Thus the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which actually matter for coexistence.
Vital rates are often measured directly in demongraphic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not yet been done.

@@

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms result from rendering one or more of $X$, $Y$, and $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present difficulties.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.
It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}. We have here used simple mathematical
and laboratory models to provide an initial exploration of the potential importance of ATAs
for coexistence. Future work should attempt to generalize lessons learned here to 
field systems with multiple species, stage structure, spatial extent, non-competitive interactions 
such as predation and facilitation that can also mediate coexistence, and other complexities.

@@

The authors thank Shyamolina Ghosh for suggesting the research problem, and Lawrence Sheppard, Max Castorani,
Jon Walter, Jude Kastens, Vadim Karatayev, Emily Arsenault, Neil Thimmesch, and Robin Snyder for helpful discussions. 
The authors were partly supported by US National
Science Foundation grants 1714195 and 2023474, the James S McDonnell Foundation, and the California Department of Fish and
Wildlife Delta Science Program. 
%***DAN: DAN: if we take Ellner off the author list add him back here, and also add him back to the acks in the README.

@@

\begin{figure}
\includegraphics[width=.38\textwidth]{../results_figs/fig6.pdf}
\caption{The contribution of ATAs to GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
the GWR itself. The ratio $\Delta_i^{[EC]}/\text{GWR}$ is shown, larger magnitudes
indicating greater importance of ATAs and values greater than $1$ corresponding to either
ATA rescue or ATA exclusion (Methods). For this system and the plotted parameters, 
$\Delta_i^{[EC]}$ was always positive, so cyan indicates exclusion of \emph{F. crotonensis} and
magenta indicates coexistence of the two diatom species. Values greater than 1 or less than -1
were remapped to 1 and -1, respectively, for readability.
For the parameters we considered, GWR of the other species, \emph{C. pseudostelligera}, was always positive both including and
excluding ATAs, so GWR for \emph{F. crotonensis} is a sufficient coexistence metric.
Standard errors of plotted quantities were always less then \Sexpr{round(se6,5)}, so are not shown.} \label{fig:fig6}
\end{figure}
